\begin{references}{99}
\bibitem{dahl72} \person{O.\@ J.\@ Dahl}, \person{E.\@ W.\@ Dijkstra}, and \person{C.\@ A.\@ R.\@ Hoare},
\emph{Structured Programming,} London, Academic Press, 1972.
\bibitem{wirth71} \person{Niklaus Wirth}, ``Program Development by Stepwise
Refinement,'' \emph{Communications of ACM,} 14, No. 4 (1971), 221-27.
\bibitem{stevens74-1} \person{W.\@ P.\@ Stevens}, \person{G.\@ J.\@ Myers}, and \person{L.\@ L.\@ Constantine},
``Structured Design,'' \emph{IBM Systems Journal,} Vol. 13, No. 2, 1974.
\bibitem{parnas72} \person{David L.\@ Parnas}, ``On the Criteria To Be Used in
Decomposing Systems into Modules,'' \emph{Communications of the ACM,}
December 1972.
\bibitem{liskov75} \person{Barbara H.\@ Liskov} and \person{Stephen N.\@ Zilles},
``Specification Techniques for Data Abstractions,'' \emph{IEEE
Transactions on Software Engineering,} March 1975.
\bibitem{parnas79} \person{David L.\@ Parnas}, ``Designing Software for Ease of
Extension and Contraction,'' \emph{IEEE Transactions on Software
Engineering,} March 1979.
\bibitem{shorre71} \person{Dewey Val Shorre}, ``Adding Modules to \Forth{},''
1980 FORML Proceedings, p. 71.
\bibitem{bern83} \person{Mark Bernstein}, ``Programming in the Laboratory,''
  unpublished paper, 1983.
\bibitem{bell72} \person{James R.\@ Bell}, ``Threaded Code,'' \emph{Communications
of ACM,} Vol. 16, No. 6, 370-72.
\bibitem{dewar} \person{Robert B.\@ K.\@ DeWar}, ``Indirect Threaded Code,''
\emph{Communications of ACM,} Vol. 18, No. 6, 331.
\bibitem{kogge82} \person{Peter M.\@ Kogge}, ``An Architectural Trail to
Threaded-Code Systems,'' \emph{Computer,} March, 1982.
\bibitem{dumse} \person{Randy Dumse}, ``The R65F11 \Forth{} Chip,'' \emph{\Forth{}
Dimensions,} Vol. 5, No. 2, p. 25.
\end{references}
